---
title: "GWAS QC: Heterozygosity"
author: "Shea Andrews"
date: "`r Sys.Date()`"
output: html_document
editor_options:
  chunk_output_type: console
---

# Heterozygosity check

Insufficient heterozygosity can indicate inbreeding or other family substructures, while excessive heterozygosity may indicate poor sample quality.

Individuals with outlying heterozygosity rates can be identified in PLINK 1.9 by typing the following command at the shell prompt:

```{bash, eval=F}
plink_1.9 --bfile raw-GWA-data \
  --extract snplist.prune.in \
  --het --out output.het
```

This produces a file containing Method-of-moments F coefficient estimates, which can be used to calculate the observed heterozygosity rate in each individual. Analysis is performed using an LD pruned snplist.

We calculate a heterozygocity similarly using observed and expected counts from the PLINK output [(Observed - Expected)/N) and exclude samples that are ± 3 sd from the cohort mean.
<br>

```{r setup heterozygosity, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

## ---- Load Pacakges ---- ##
library(readr)
library(dplyr)
library(tidyr)
library(ggplot2)
library(knitr)
library(DT)
tabcap <- function (name, caption) {
    sprintf("<table width=\"100%%\"><caption>(#tab:%s) %s </caption></table>", name, caption)
}
disptab <- function(x) {
    if (knitr::opts_knit$get("rmarkdown.pandoc.to") == "html") {
          dtopts <- list(dom = 'Bfrtip', buttons = c('copy', 'csv', 'excel', 'pdf', 'print'))
    return(datatable(x, extensions = 'Buttons', options = dtopts))
      } else {
            return(kable(x))
      }
}
```

```{r read_data heterozygosity, echo=F, cache=T}
## ---- Read in Data ----##

het.raw <- as_tibble(read.table(snakemake@input[["HetFile"]], header = TRUE, check.names = FALSE, as.is = TRUE))
```

```{r data_wrangling heterozygosity, echo=F}
## caluclate heterozygosity
het <- het.raw %>%
  rename(O = `O(HOM)`, E = `E(HOM)`, N = `N(NM)`) %>%
  mutate(Het = (N - O) / N)

##  Calculate exclusion thresholds
upper.het <- mean(het$Het) + sd(het$Het)*3
lower.het <- mean(het$Het) - sd(het$Het)*3

##  Exclusion of samples
het <- het %>%
  mutate(exclude = ifelse(Het >= upper.het | Het <= lower.het, TRUE, FALSE))

exclude.samples <- het %>% filter(exclude == TRUE)
```

Figure \@ref(fig:plothet) displays the distrubution of heterozygosity in `r snakemake@wildcards[["sample"]]`. Samples with excessive (Het > `r upper.het`) or deficient (Het < `r lower.het`) heterozygosity are colored red. Table \@ref(tab:het) displays samples that are to be excluded.
```{r plothet, echo=FALSE, fig.width=7.5, fig.height=4.7, fig.align='center', fig.cap='Distribution of heterozygosity'}
ggplot(het, aes(x = Het, fill = exclude)) + geom_histogram(binwidth = 0.001) +
  geom_vline(xintercept = upper.het, colour = 'red', linetype = 2) +
  geom_vline(xintercept = lower.het, colour = 'red', linetype = 2) +
  theme_bw() + scale_fill_manual(values = c("#377EB8", "#E41A1C")) +
  theme(legend.position = 'bottom') +
  labs(x = 'Heterozygosity')
```
\

`r tabcap("het", "Samples with outlying heterozygosity rates")`
```{r table heterozygosity, echo=FALSE}
##  Print samples with outlying heterozygosity
exclude.samples %>% disptab
```

```{r cleanup_het, include=F}
rm(exclude.samples, het, het.raw, upper.het, lower.het)
```
