```{r prepLabels, include=FALSE, cache=F}
library(dplyr)
library(readr)
library(tibble)
library(tidyr)
library(purrr)
library(ggplot2)
library(cowplot)
library(ggtern)
library(colorspace)
library(plotly)
library(here)
library(kableExtra)

tg_subpop <- "workflow/resources/tg_subpops.tsv"

if (!file.exists(here(tg_subpop))) here <- function(x) paste0(snakmake@params[["rwd"]], "/", x)

tg_labs <- tg_subpop %>%
  here() %>%
  read_tsv(col_types = cols(.default = "c")) %>%
  arrange(spop, pop)

sp_labs <- tg_labs %>% select(spop, description_spop) %>% distinct %>% deframe

paste_pretty <- function(x) {
  begin <- paste(x[1:length(x) - 1], collapse = ", ")
  sprintf("%s and %s", begin, x[[length(x)]])
}

sp_desc <- tg_labs %>%
  mutate(desc = sprintf("%s (%s)", description_pop, pop)) %>%
  select(spop, desc) %>%
  chop(desc) %>%
  mutate(desc = sapply(desc, paste_pretty))


```

### `r sp_labs[snakemake@params[["superpop"]]]` Population Structure

Populations have been inferred based on minimum distance from the geometric median of the first 10 principle components.

```{r pcastats, echo=F}
tab_1 %>%
  mutate(across(where(is.double), ~ round(.x, digits = 3))) %>%
  kable(booktabs = T, caption = "Geometric medians of each PC in reference superpopulations") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

The following plots show the population structure of `r snakemake@wildcards[["sample"]]` (Black Triangles) based on the first three principal components. In Fig. \@ref(fig:ternplots_assign), the reference and projected samples are displayed on a ternary plot. Fig. \@ref(fig:3PCinteractiveSuperPops) contains the same data as an interactive 3d plot.

```{r ternplots_assign, echo=F, fig.width=7.5, fig.height=3.4, fig.align='center', fig.cap='Ancestry assignment viewed on PCs 1-3'}


# Format data for ploting ternery PCA
dat_tern <- select(pca, PC1, PC2, PC3) %>%
  mutate(across(starts_with("PC"), ~ . - min(.))) %>%
  as.matrix() %>%
  prop.table(1)  %>%
  as_tibble() %>%
  bind_cols(select(pca, IID, cohort, superpop_infered)) %>%
  mutate(cohort2 = ifelse(cohort == "Reference", "Reference", "Sample"))

mark_refs <- . %>%
  mutate(superpop_infered = ifelse(
    cohort == "Reference",
    sprintf("%s (ref)", superpop_infered),
    superpop_infered))

pc_ternplot <- function(x) {
  ggtern(x, aes(x=PC1, y=PC2, z=PC3, color = superpop_infered, shape = cohort2)) +
    geom_point(size = 1) +
    scale_shape_manual(values = c(Reference = 16, Sample = 17)) +
    theme_bw() +
    theme_showarrows() +
    theme_notitles() +
    theme_hidelabels() +
    theme(text = element_text(size=10))
}

pca_col2 <- pca_col %>% filter(color != "black") %>% select(-n)
pca_col2 <- pca_col2 %>%
  mutate(superpop = sprintf("%s (ref)", superpop),
         color = lighten(desaturate(color, amount = 0.3), amount = 0.8)) %>%
  bind_rows(pca_col2) %>%
    deframe %>%
  as.list()

p_ref <- dat_tern %>%
  filter(cohort == "Reference") %>%
  pc_ternplot +
    scale_color_manual(values = pca_col2[!grepl("ref", names(pca_col2))]) +
    labs(color = "Population")

## Overlay sample on reference population pca space

p_sample_dat <- dat_tern %>% mark_refs

p_sample <- p_sample_dat %>%
  pc_ternplot +
    scale_color_manual(values = pca_col2) +
    theme(legend.position = "none")

legend <- get_legend(
  # create some space to the left of the legend
  ggplotGrob(p_ref + theme(legend.box.margin = margin(0, 0, 0, 12), legend.title = element_blank()))
)

p1 <- plot_grid(
  ggplotGrob(p_ref + theme(legend.position="none", plot.margin=unit(c(-15,-7,-7,-7), "mm"))),
  ggplotGrob(p_sample + theme(plot.margin=unit(c(-15,-7,-7,-7), "mm"))),
  legend,
  labels = c('Reference', 'Sample'), label_size = 12, ncol = 3,
  rel_widths = c(1, 1, 0.4)
  )

ggsave("plot.pdf", plot = p1, width = 7.5, height = 3.4, unit = "in")

p1

detach("package:ggtern", unload=TRUE)
```

```{r umap, echo=F, include=F, eval=F, fig.width=7.5, fig.height=3.4, fig.align='center', fig.cap='Ancestry assignment viewed on PCs 1-3'}
# pca_umap <- select(pca, starts_with("PC")) %>%
#   umap() %$%
#   set_colnames(layout, c("UMPC1", "UMPC2")) %>%
#   as_tibble() %>%
#   bind_cols(pca, .) %>%
#   mutate(cohort2 = ifelse(cohort == "Reference", "Reference", "Sample"))
#
# pca_umap %>%
#   mark_refs %>%
#   ggplot(aes(x=UMPC1, y=UMPC2, color = superpop_infered, shape = cohort2)) +
#   geom_point(size = 1) +
#   scale_shape_manual(values = c(Reference = 16, Sample = 17)) +
#   theme_bw() +
#   theme(text = element_text(size=10), legend.box.margin = margin(0, 0, 0, 12), legend.title = element_blank()) +
#   scale_color_manual(values = pca_col2)
```

The following plots are interactive so that outliers can be identified graphically for further removal. Geometric medians are applied across all `r as.character(nrow(eigenval))` principle componants, so some of the clusters used to distinguish ancestries may not be visible on the plot.

```{r 3PCinteractiveSuperPops_tern, echo=F, fig.width=7.5, fig.height=5, fig.align='center', fig.cap='Tridimensional plot of ancestery clustering of PC 1, 2 & 3'}

# Plotly drops colors with only one point
fix_plotly_drops <- function(df) {
  singletons <- df %>%
    group_by(superpop_infered) %>%
    filter(n() == 1) %>%
    ungroup
  bind_rows(df, singletons)
}

axis <- function(title) {
  list(
    title = title,
    titlefont = list(
      size = 15
    ),
    tickfont = list(
      size = 8
    ),
    tickcolor = 'rgba(0,0,0,0)',
    ticklen = 5
  )
}

p_sample_dat_plotly <- p_sample_dat %>% fix_plotly_drops

plot_tern_interact <- plot_ly() %>%
  config(displaylogo = FALSE) %>%
  add_trace(data = p_sample_dat_plotly %>% filter(cohort2 == "Reference"),
            a = ~PC2, b = ~PC1, c = ~PC3, color = ~superpop_infered,
    type = 'scatterternary', mode = 'markers', marker = list(size = 3),
    colors = unlist(pca_col2), hoverinfo="none") %>%
  add_trace(data = p_sample_dat_plotly %>% filter(cohort2 != "Reference"),
            a = ~PC2, b = ~PC1, c = ~PC3, color = ~superpop_infered,
    type = 'scatterternary', mode = 'markers', marker = list(size = 3),
    colors = unlist(pca_col2), hoverinfo = 'text',
    text = ~paste('</br> ID: ', IID, '</br> Population: ', superpop_infered,
                '</br> PC1: ', round(PC1, 2), '</br> PC2: ', round(PC2, 2),
                '</br> PC3: ', round(PC3, 2)))

plot_tern_interact %>%
    plotly::layout(
      ternary = list(
        sum = 100,
        aaxis = axis('PC 2'),
        baxis = axis('PC 1'),
        caxis = axis('PC 3')
      )
    )


```

A 3D version follows for further exploration.

```{r 3PCinteractiveSuperPops, echo=F, fig.width=7.5, fig.height=4.7, fig.align='center', fig.cap='Tridimensional plot of ancestery clustering of PC 1, 2 & 3'}

##  Plot superpops, PC1 + PC2 + PC3
plot_ly(pca %>% mark_refs %>% fix_plotly_drops,
        x = ~PC2, y = ~PC1, z = ~PC3, color = ~superpop_infered,
  type = 'scatter3d', mode = 'markers', marker = list(size = 3),
  colors = unlist(pca_col2), hoverinfo = 'text',
  text = ~paste('</br> ID: ', IID, '</br> Population: ', superpop_infered,
                '</br> PC1: ', round(PC1, 2), '</br> PC2: ', round(PC2, 2),
                '</br> PC3: ', round(PC3, 2)))

```



### Population Outliers

```{r early_exit, echo=F, results='asis', eval=no_outliers}
cat('No outliers were removed')
knitr::knit_exit()
```

Individuals not in the selected ancestry(s) based on the above inference were removed.

`r tabcap("popoutliers", "Population outliers")`

```{r, population outliers table, echo=F, results='asis'}
tab_pop_exclusions %>%
  disptab
```
