```{r prepLabels, include=FALSE, cache=F}
library(dplyr)
library(readr)
library(tibble)
library(tidyr)
library(purrr)

tg_subpop <- "workflow/resources/tg_subpops.tsv"

if (!file.exists(here(tg_subpop))) here <- function(x) paste0(rwd, "/", x)

tg_labs <- tg_subpop %>%
  read_tsv(col_types = cols(.default = "c")) %>%
  arrange(spop, pop)

sp_labs <- tg_labs %>% select(spop, description_spop) %>% distinct %>% deframe

paste_pretty <- function(x) {
  begin <- paste(x[1:length(x) - 1], collapse = ", ")
  sprintf("%s and %s", begin, x[[length(x)]])
}

sp_desc <- tg_labs %>%
  mutate(desc = sprintf("%s (%s)", description_pop, pop)) %>%
  select(spop, desc) %>%
  chop(desc) %>%
  mutate(desc = sapply(desc, paste_pretty)) %>%
  deframe
```

### `r sp_labs[params$superpop]` Population Structure

The following plots show the population structure of `r params$Sample` (Black Triangles) based on the first two (Fig. \@ref(fig:2PCstaticSuperPops)) and three (Fig. \@ref(fig:3PCinteractiveSuperPops)) principal components compared with the reference populations from 1000 Genomes. The static plot is zoomed into the `r sp_labs[params$superpop]` refernce super population composed of `r sp_desc[params$superpop]`.

```{r 2PCstaticSuperPops, echo=F, fig.width=7.5, fig.height=4.7, fig.align='center', fig.cap='Ancestery clustering based on PC 1 & 2'}
##  Plot superpops, PC1 + PC2
ggplot(pca, aes(x = PC2, y = PC1, color = superpop, shape = cohort)) +
  geom_point() + facet_zoom(xy = superpop == params$superpop) +
  scale_color_manual(values = pca_col$color) + coord_fixed(ratio = 1) +
  theme_bw() + theme(legend.position = 'right')
```

```{r 3PCinteractiveSuperPops, echo=F, fig.width=7.5, fig.height=4.7, fig.align='center', fig.cap='Tridimensional plot of ancestery clustering of PC 1, 2 & 3'}

##  Plot superpops, PC1 + PC2 + PC3
plot_ly(pca, x = ~PC2, y = ~PC1, z = ~PC3, color = ~superpop,
  type = 'scatter3d', mode = 'markers', marker = list(size = 3),
  colors = pca_col$color, hoverinfo = 'text',
  text = ~paste('</br> ID: ', IID, '</br> Population: ', Population,
                '</br> PC1: ', round(PC1, 2), '</br> PC2: ', round(PC2, 2),
                '</br> PC3: ', round(PC3, 2)))
```

### Population Outliers

Individuals of non-`r sp_labs[params$superpop]` ancestry were identified by determining the mean and standard deviation of the 10 principal components scores for the `r params$superpop` super-population (Table 1). Participants in `r params$Sample` who were \pm `r as.character(sdev)` SD from the `r params$superpop` population mean were determined to be of non-`r sp_labs[params$superpop]` ancestry (Tab. \@ref(tab:pcastats)).

```{r pcastats, echo=F}
tab_1 %>% kable(booktabs = T, caption = sprintf("Mean and SD of PCs in %s population", params$superpop)) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```


```{r early_exit, echo=F, results='asis', eval=no_outliers}
cat('Figures and Table for population outliers are omitted because there are no outliers.')
knitr::knit_exit()
```

Fig. \@ref(fig:popOutliers2PC) and Fig. \@ref(fig:popOutliers3PC) display the population structure of `r params$Sample` with `r params$superpop` population outliers highlighted in red. Tab. \@ref(tab:popoutliers) lists all population outliers for removal.

```{r, popOutliers2PC, echo=F, fig.width=7.5, fig.height=4.7, fig.align='center', fig.cap=sprintf('%s population outliers on PC 1 & 2', params$superpop)}
ggplot(pca_sample, aes(x = PC2, y = PC1, color = pop_outliers)) +
  geom_point() + coord_fixed(ratio = 1) + theme_bw() +
  facet_zoom(xy = superpop %in% c(params$superpop) | pop_outliers == F) +
  scale_color_manual(name = c('Population Outlier'),
                     values = c("#377EB8", "#E41A1C"))
```

```{r, popOutliers3PC, echo=F, fig.width=7.5, fig.height=4.7, fig.align='center', fig.cap='Tridimensional plot displaing population outliers on PC 1, 2, & 3'}
plot_ly(pca_sample, x = ~PC2, y = ~PC1, z = ~PC3, color = ~pop_outliers,
        type = 'scatter3d', mode = 'markers', marker = list(size = 3),
        colors = c("#377EB8", "#E41A1C"), hoverinfo = 'text',
        text = ~paste('</br> ID: ', IID,
                      '</br> PC1: ', round(PC1, 2),
                      '</br> PC2: ', round(PC2, 2),
                      '</br> PC3: ', round(PC3, 2)))
```


`r tabcap("popoutliers", sprintf("%s Population Outliers for each PC", params$superpop))`

```{r, population outliers table, echo=F, results='asis'}
tab_pop_exclusions %>%
  disptab
```
