---
title: "Sex inconsistencies"
author: "Shea Andrews"
date: "`r Sys.Date()`"
output:
  html_document:
    df_print: paged
params:
  Path_SexFile:
  rwd:
editor_options:
  chunk_output_type: console
---

# Sex Discordance

Samples with discordance between self-reported and genetically predicted sex likely have errors in sample handling, such as sample swaps. Predicted sex can be determined by calculating X chromosome heterozygosity using an F test, because biological men have one X chromosome and women have two. An F value of ~0.99 indicates males, and an F value of ~0.03 indicates females. Furthermore, checking X chromosome heterozygosity may reveal sex chromosome anomalies (~0.28 in reported females; ~0.35 in males).

Since sex discordance may be due to sample swaps or to incorrect phenotyping, sex discordant samples should generally be removed unless a swap can be reliably resolved.

Identification of individuals with discordent sex can be done in PLINK 1.9 by typing the following at the shell prompt, which will produce a list of individuals with discordent sex data.


```{bash, eval=F}
plink_1.9 --bfile raw-GWA-data  \
  --check-sex --out --out output.sexcheck
```
<br>

```{r setup sexcheck, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = params$rwd)

## ---- Load Pacakges ---- ##
library(readr)
library(dplyr)
library(tidyr)
library(ggplot2)
library(knitr)
library(DT)
tabcap <- function (name, caption) {
  sprintf("<table width=\"100%%\"><caption>(#tab:%s) %s </caption></table>", name, caption)
}
disptab <- function(x) {
  if (knitr::opts_knit$get("rmarkdown.pandoc.to") == "html") {
    dtopts <- list(dom = 'Bfrtip', buttons = c('copy', 'csv', 'excel', 'pdf', 'print'))
    return(datatable(x, extensions = 'Buttons', options = dtopts))
  } else {
    return(kable(x))
  }
}
```

```{r read in sex check data, echo=F, message=F, warning=F}
## ---- Read in Data ----##

sexcheck.raw <- read_table2(params$Path_SexFile)

##  recode sex varibles
sexcheck <- sexcheck.raw %>%
  mutate(PEDSEX = recode(PEDSEX, '2' = 'Female', '1' = 'Male'))

##  Exclude samples with no sex inconsistencies
exclude.samples <- sexcheck %>%
  filter(STATUS == 'PROBLEM') %>%
  mutate(PEDSEX = recode(PEDSEX, '2' = 'Female', '1' = 'Male'))
```

The following plot (Fig. \@ref(fig:sexplot)) displays the X Chromosome heterozygosity for self reported sex, with samples with problems highlighted in red. Table \@ref(tab:sextab) displays the individule records that should be excluded from further downstream analysis.
```{r sexplot, echo=FALSE, fig.width=5.5, fig.height=4.7, fig.align='center', fig.cap='X chromosome heterozygosity in self reported males & females'}
ggplot(data = sexcheck, aes(x = as.factor(PEDSEX), y = F, colour = STATUS, shape = STATUS)) +
  geom_jitter() +
  scale_color_manual(values = c( "#377EB8", "#E41A1C")) +
  theme_bw() + labs(x = 'Self reported sex', y = 'X CHR Heterozygocity (F)') + theme(legend.position="bottom")
```
<br>

`r tabcap("sextab", "Samples with sex discordance")`

```{r sextab, echo=FALSE}
##  Print samples with sex inconsistencies
exclude.samples %>% disptab

```


```{r write_out, echo=FALSE}

##  write out samples to be excluded
#exclude.samples <- exclude.samples %>%  select(FID, IID)
#write_tsv(exclude.samples, outfile, col_names = T)

```
<br>
