---
title: "GWAS QC: Heterozygosity"
author: "Shea Andrews"
date: "`r Sys.Date()`"
output: html_document
params:
    Path_HetFile:
    Sample:
    rwd:
editor_options:
  chunk_output_type: console
---

## Heterozygosity check

Samples with defficint heterozygosity can indicate inbreeding or other substructures in that person individule, while samples with excessive heterozyosity may indicate poor sample quality. Samples that are either ± 3 sd from the group mean are excluded.


Identifing individules with outlying heterozygocity rates can be done in PLINK by typing the following command at the shell prompt. This produce a file containing the Method-of-moments F coefficient estimates, which can be used to calculate the observed heterozygosity rate in each individule. Analysis should be performed using an LD pruned snplist.

```{bash, eval=F}
plink_1.9 --bfile raw-GWA-data \
  --extract snplist.prune.in \
  --het --out output.het
```

<br>

```{r setup heterozygosity, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = params$rwd)

## ---- Load Pacakges ---- ##
library(tidyverse)
library(ggplot2)
library(knitr)

```

```{r read_data heterozygosity, echo=F, cache=T}
## ---- Read in Data ----##

het.raw <- as.tibble(read.table(params$Path_HetFile, header = TRUE, check.names = FALSE, as.is = TRUE))
```

```{r data_wrangling heterozygositcy, echo=F}
## caluclate heterozygosity
het <- het.raw %>%
  rename(O = `O(HOM)`, E = `E(HOM)`, N = `N(NM)`) %>%
  mutate(Het = (N - O) / N)

##  Calculate exclusion thresholds
upper.het <- mean(het$Het) + sd(het$Het)*3
lower.het <- mean(het$Het) - sd(het$Het)*3

##  Exclusion of samples
het <- het %>%
  mutate(exclude = ifelse(Het >= upper.het | Het <= lower.het, TRUE, FALSE))

exclude.samples <- het %>% filter(exclude == TRUE)
```

Figure 1 displays the distrubution of heterozygocity in `r params$Sample`. Samples with excessive (Het > `r upper.het`) or deficient (Het < `r lower.het`) Heterozygocity are colour red. Table 1 displays the sample that are to be excluded.
```{r plot heterozygocity, echo=FALSE, fig.width=7.5, fig.height=4.7, fig.align='center', fig.cap='Fig. 1: Distribution of heterozygocity'}
ggplot(het, aes(x = Het, fill = exclude)) + geom_histogram(binwidth = 0.001) +
  geom_vline(xintercept = upper.het, colour = 'red', linetype = 2) +
  geom_vline(xintercept = lower.het, colour = 'red', linetype = 2) +
  theme_bw() + scale_fill_manual(values = c("#377EB8", "#E41A1C")) +
  theme(legend.position = 'bottom') +
  labs(x = 'Heterozygosity')
```
\

***Table 1:*** Samples with outlying heterozygocity rates
```{r table heterozygosity, echo=FALSE}
##  Print samples with sex inconsistencies
as.data.frame(exclude.samples)
```


<br>
